---
title: "Categorical spatial data extraction around buffered points in R"
description: |
  Computing the proportion of land cover types using R and the raster package.
author:
  - name: Maxwell B. Joseph
    url: {}
date: 2014-11-08
output:
  distill::distill_article:
    self_contained: false
editor_options: 
  chunk_output_type: console
categories: 
  - spatial
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Quantifying categorical spatial data (e.g. land cover) around points can be done 
in a variety of ways, some of which require considerable amounts of patience, 
clicking around, and/or cash for a license.
Here's a bit of code that I cobbled together to quickly extract land cover data
from the [National Land Cover Database](http://www.mrlc.gov/index.php) for 
a buffered point between Denver and Boulder - two cities in the state of 
Colorado.

First, get data using the FedData package:

```{r get-data}
library(knitr)
library(raster)
library(FedData)
library(rgdal)

site_df <- data.frame(city = c('Denver', 'Boulder'), 
                      lat = c(39.7392, 40.0150), 
                      lon = c(-104.9903, -105.2705))
midpoint <- data.frame(lat = mean(site_df$lat), 
                       lon = mean(site_df$lon))

# create spatial point data frame
coordinates(site_df) <- ~lon + lat
proj4string(site_df) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84")

coordinates(midpoint) <- ~lon + lat
proj4string(midpoint) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84")

nlcd_raster <- get_nlcd(site_df, 
                        label = 'categorial-extraction', 
                        year = 2011, 
                        raw.dir = '.', 
                        extraction.dir = '.')

# reproject midpoint to raster's crs
midpoint <- spTransform(midpoint, projection(nlcd_raster))

buffer_distance_meters <- 5000

# visualize buffered point and land cover data
buff_shp <- buffer(midpoint, buffer_distance_meters)
plot(buff_shp) # I will plot over this, but it sets the plotting extent
plot(nlcd_raster, add = TRUE)
plot(buff_shp, add = TRUE)
```

Now, we can use `raster::extract` to extract raster data around our point
within some buffer, specified in meters. 

```{r extract-data}
landcover <- extract(nlcd_raster, midpoint, buffer = buffer_distance_meters)
```

But this object does not immediately provide the proportions of each cover type. 
Instead, it contains values from the cells within the buffer:

```{r inspect-landcover}
str(landcover)
```

We can get the proportions as follows:

```{r get-proportions}
landcover_proportions <- lapply(landcover, function(x) {
  counts_x <- table(x)
  proportions_x <- prop.table(counts_x)
  sort(proportions_x)
  })
str(landcover_proportions)
```

To make this data more useful we can add land cover class names, which are 
stored as data attributes (the data are stored as integers, but we want
character class names).

```{r}
# get a data frame to map integer cover values to classes
nlcd_cover_df <- subset(nlcd_raster@data@attributes[[1]], 
                        NLCD.2011.Land.Cover.Class != "")
kable(nlcd_cover_df)
```

Now we can use this data frame to map the cover type integers in 
`landcover_proportions` to character descriptions of the cover types. 

```{r}
lapply(landcover_proportions, function(x) {
  current_int_names <- names(x)
  label_indices <- match(current_int_names, as.character(nlcd_cover_df$Value))
  new_char_names <- nlcd_cover_df$NLCD.2011.Land.Cover.Class[label_indices]
  names(x) <- new_char_names
  x
})
```

**Resources**

- [Large .img file processing in R (GIS) on Stack Overflow by Israel Del Toro](http://stackoverflow.com/questions/15824853/large-img-file-processing-in-r-gis)
- [NLCD website](http://www.mrlc.gov/index.php)
