---
title: "First year books"
description: |
  10 books I wish I had entering graduate school.
author:
  - name: Maxwell B. Joseph
    url: {}
date: 2015-09-08
output:
  distill::distill_article:
    self_contained: false
categories: 
  - rants
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

I had to read a lot of books in graduate school.
Some were life-changing, and others were forgettable.

If I could bring a reading list back in time for my 'first year' graduate self, it would include the following:

### [Bayesian Data Analysis](https://www.crcpress.com/Bayesian-Data-Analysis-Third-Edition/Gelman-Carlin-Stern-Dunson-Vehtari-Rubin/9781439840955)

Third Edition, by Andrew Gelman, John B. Carlin, Hal S. Stern, David B. Dunson, Aki Vehtari, and Donald B. Rubin

Probably the most useful book I've ever owned. Has staying power - can be used
as a reference.

### [The Art of R Programming](https://www.nostarch.com/artofr.htm)

by Norman Matloff

This book made me less bad at programming in R early on.

### [Causality](http://www.cambridge.org/us/catalogue/catalogue.asp?isbn=9780521895606)

Second Edition, by Judea Pearl

Ecology is complicated. We often lack replicated controlled experiments with 
random treatment assignment. This book helped me organize my thinking around
how to translate mechanistic knowledge to statistical models. 

### [Statistics for Spatio-Temporal Data](http://www.wiley.com/WileyCDA/WileyTitle/productCd-EHEP002348.html)

by Noel Cressie and Christopher Wikle

A thoughtful treatment of hierarchical modeling in a spatial, temporal, and 
spatiotemporal context. Has breadth with a healthy dose of outside references
for depth. My dog ate this one, but it was great to have.

### [Ecological Models and Data in R](http://press.princeton.edu/titles/8709.html)

by Benjamin M. Bolker

Covers fundamental ideas about likelihood and process-oriented modeling while 
building R proficiency.

### [Bayesian Models: A Statistical Primer for Ecologists](http://press.princeton.edu/titles/10523.html)

by N. Thompson Hobbs & Mevin B. Hooten

An introduction to the process of model building and estimation for 
non-math/stats oriented readers. Thoughtful treatment of notation, helped me to
better understand how to communicate models. 

### [Data Analysis Using Regression and Multilevel/Hierarchical Models](http://www.cambridge.org/us/academic/subjects/statistics-probability/statistical-theory-and-methods/data-analysis-using-regression-and-multilevelhierarchical-models)

by Andrew Gelman and Jennifer Hill

A gentle introduction to multilevel modeling, with plenty of graphics and 
integration with R.

### [Statistical Inference](http://www.amazon.com/Statistical-Inference-George-Casella/dp/0534243126)

Second Edition, by George Casella and Roger L. Berger

Essential for understanding the mathematical and probabilistic foundations of 
statistics. Read it after brushing up on calculus. Checked this out from the 
library, and my dog ate it. Had to buy a copy to replace, which was not cheap. 
But, later found a pdf online. 

### [Linear Algebra](http://store.doverpublications.com/048663518x.html)
by George Shilov

I wish I had taken a class in linear algebra as an undergraduate, but I instead
had to catch up in my first year of grad school. This book made it relatively 
painless. Literally found it on the shelf of the office I moved into on day one.

### [Single and Multivariable Calculus](https://www.whitman.edu/mathematics/calculus_online/)
by David Guichard and friends

Because I took a few calculus classes in high school and college and didn't know 
why.

### [Mathematical Tools for Understanding Infectious Disease Dynamics](http://press.princeton.edu/titles/9916.html)

by Odo Diekmann, Hans Heesterbeek & Tom Britton

Mathematical epidemiology is a huge topic. This book introduces common models 
and approaches from first principles, with plenty of problems along the way to 
make sure you're following along. Read it with a notebook and pencil handy.


```{r plot-tape, fig.cap="Credit: Wellcome Library, London. Wellcome Images images@wellcome.ac.uk http://wellcomeimages.org Plague treatise - a full page woodcut representing a man at a table studying a book, he is surrounded by four persons, - one holding a urine glass, Woodcut Darin durch sechs kurtzer Buchlin vil Heimlichkeiten der Natur beschriben werden Albertus Magnus Published: 1551 ", fig.width = 10, fig.height = 12}
library(imager)
im <- load.image("https://upload.wikimedia.org/wikipedia/commons/3/3d/A._Magnus%2C_%22Darin_durch_sechs...%22%2C_1551%3B_man_studying_a_book_Wellcome_L0003173.jpg")
plot(im, axes = FALSE)
```