---
title: "Bayesian model II regression in JAGS"
description: |
  Fitting a regression model with uncertainty in the explanatory variable.
author:
  - name: Maxwell B. Joseph
    url: {}
date: 05-27-2013
output:
  radix::radix_article:
    self_contained: false
categories: 
  - jags
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Regression is a mainstay of ecological and evolutionary data analysis. For 
example, a disease ecologist may use body size (e.g. a weight from a scale with
measurement error) to predict infection. Classical linear regression assumes no 
error in covariates; they are known exactly. This is rarely the case in ecology, 
and ignoring error in covariates can bias regression coefficient estimates. This 
is where model II (aka errors-in variables and measurement errors) regression 
models come in handy. Here I'll demonstrate how to construct such a model in a 
Bayesian framework, where substantive prior knowledge of covariate error 
facilitates less-biased parameter estimates.

Here's a quick illustration of the problem: I'll generate data from a known 
simple linear regression model, and fit models that ignore or incorporate error
in the covariate.

```{r simulate-data}
library(rjags)
library(ggmcmc)

# simulate covariate data
n <- 40
sdx <- 6
sdobs <- 5
taux <- 1 / (sdobs * sdobs)
truex <- rnorm(n, 0, sdx)
errorx <- rnorm(n, 0, sdobs)
obsx <- truex + errorx

# simulate response data
alpha <- 0
beta <- 10
sdy <- 20
errory <- rnorm(n, 0, sdy)
obsy <- alpha + beta*truex + errory
observed_data <- data.frame(obsx = obsx, obsy = obsy)
parms <- data.frame(alpha, beta)
```

Ignoring error in the covariate:

```{r ignore-covariate-error}
# bundle data
jags_d <- list(x = obsx, y = obsy, n = length(obsx))

# write model
cat("
    model{
## Priors
alpha ~ dnorm(0, .001)
beta ~ dnorm(0, .001)
sdy ~ dunif(0, 100)
tauy <- 1 / (sdy * sdy)

## Likelihood
  for (i in 1:n){
    mu[i] <- alpha + beta * x[i]
    y[i] ~ dnorm(mu[i], tauy)
  }
}
",
    fill=TRUE, file="yerror.txt")

# initiate model
mod1 <- jags.model("yerror.txt", data=jags_d,
                   n.chains=3, n.adapt=1000)

# simulate posterior
out <- coda.samples(mod1, n.iter=1000, thin=1,
                    variable.names=c("alpha", "beta", "sdy"))

# store parameter estimates
ggd <- ggs(out)
a <- ggd$value[which(ggd$Parameter == "alpha")]
b <- ggd$value[which(ggd$Parameter == "beta")]
d <- data.frame(a, b)
```

Incorporating error in the covariate: I'm assuming that we have substantive 
knowledge about covariate measurement represented in the prior for the precision 
in X. Further, the prior for the true X values reflects  knowledge of the 
distribution of our X value in the population from which the sample was taken.

```{r deal-with-error}
# specify model
cat("
    model {
## Priors
alpha ~ dnorm(0, .001)
beta ~ dnorm(0, .001)
sdy ~ dunif(0, 100)
tauy <- 1 / (sdy * sdy)
taux ~ dunif(.03, .05)

## Likelihood
  for (i in 1:n){
    truex[i] ~ dnorm(0, .04)
    x[i] ~ dnorm(truex[i], taux)
    y[i] ~ dnorm(mu[i], tauy)
    mu[i] <- alpha + beta * truex[i]
  }
}
    ", fill=T, file="xyerror.txt")

# bundle data
jags_d <- list(x = obsx, y = obsy, n = length(obsx))

# initiate model
mod2 <- jags.model("xyerror.txt", data=jags_d,
                   n.chains=3, n.adapt=1000)

# simulate posterior
out <- coda.samples(mod2, n.iter=30000, thin=30,
                    variable.names=c("alpha", "beta", "tauy", "taux"))
# store parameter estimates
ggd <- ggs(out)
a2 <- ggd$value[which(ggd$Parameter == "alpha")]
b2 <- ggd$value[which(ggd$Parameter == "beta")]
d2 <- data.frame(a2, b2)
```

Now let's see how the two models perform.

```{r compare-models}
ggplot(observed_data, aes(x=obsx, obsy)) +
  geom_abline(aes(intercept=a, slope=b), data=d, 
              color="red", alpha=0.05) +
  geom_abline(aes(intercept=a2, slope=b2), data=d2, 
              color="dodgerblue", alpha=0.05) +
  geom_abline(aes(intercept=alpha, slope=beta),
              data=parms, color="green", size=1.5, linetype="dashed") +
  geom_point(shape=19, size=3) +
  theme_minimal() +
  xlab("X values") + 
  ylab("Observed Y values") +
  ggtitle("Model results with and without modeling error in X")
```

The dashed green line shows the model that generated the data, i.e. the "true" 
line. The red lines show the posterior for the naive model ignoring error in X, 
while the less-biased blue lines show the posterior for the model incorporating 
error in X.
