---
title: "Yes, but does it (still) run?"
description: |
  Migrating from Jekyll to radix, with some reflections on the past 6 years.
author:
  - name: Maxwell B. Joseph
    url: {}
date: 12-30-2018
output:
  radix::radix_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

I haven't blogged recently.
It was useful as a PhD student to wrap my head around new 
methods and track my path from a code-naive field ecologist to a 
not entirely incompetent R programmer in 2016 when I graduated. 
But, much has changed in the 5-6 years since I began the blog. 
I've developed new skills, and the tools around R programming have matured.
It's easier now to publish content with R markdown
with tools like [blogdown](https://github.com/rstudio/blogdown) and 
[radix](https://github.com/rstudio/radix) than it was six years ago. 

My old workflow involved writing posts in R markdown, then generating markdown
and publishing with a static site generator (first
[Octopress](https://github.com/imathis/octopress), then
[Jekyll](https://github.com/jekyll/jekyll)). 
This is fine, but I ended up losing track of most of the original R markdown
files that generated the markdown being served on the site.

Looking back on 2018, I've come to better appreciate 
continuous integration through Travis CI, CircleCI, and AppVeyor.
It's nice to know when builds break and code stops working. 
As I was taking stock of the past year, I realized that I did not know whether 
the code that I had posted years ago still worked, and I didn't have the .Rmd
files anymore to find out.

```{r plot-img, fig.width=10, fig.height = 7, fig.cap = "An abandoned car that may have been useful in its time, but no longer runs."}
library(imager)
im <- load.image("https://upload.wikimedia.org/wikipedia/commons/thumb/5/57/Abandoned_Car_%28155561965%29.jpeg/1280px-Abandoned_Car_%28155561965%29.jpeg")
plot(im, axes = FALSE)
```

This bothered me. 
First, I didn't want to be responsible for publishing code that doesn't run. 
Second, I also realized that years ago, I did not appreciate
portability as much as I do today, e.g., 
when I felt comfortable assuming that users would need to manually download
data from a website and store it in a specific directory on their filesystem 
to run some code. 
With time off over the holidays, I figured I would migrate my old site from 
Jekyll to radix and put the resulting site under continuous integration to 
be sure that the answer to the question "does it (still) run?" is yes.

The result is this site (as of Dec 2018), and unsurprisingly, not all of the 
code that I had posted previously still ran. 
But, now with radix I have 
[Travis CI building the posts](https://travis-ci.org/mbjoseph/mbjoseph.github.io) 
regularly, so at least if something breaks in the future, I should know sooner.
