---
title: "Dynamic occupancy models in Stan"
description: |
  Dynamic multi-year occupancy models, marginalizing over latent occurrence states.
author:
  - name: Maxwell B. Joseph
    url: {}
date: 2014-11-14
output:
  radix::radix_article:
    self_contained: false
bibliography: biblio.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Occupancy modeling is possible in Stan 
[as shown here](https://groups.google.com/forum/#!topic/stan-users/ZuLmNl4Q9xc), 
despite the lack of support for integer parameters (without marginalization).
In many Bayesian applications of occupancy modeling, the true occupancy states 
(0 or 1) are directly modeled, but this can be avoided by marginalizing out the 
true occupancy state.
The Stan [manual](http://mc-stan.org/manual.html) (pg. 96) gives an example of 
this kind of marginalization for a discrete change-point model.

Here's a Stan implementation of a dynamic (multi-year) occupancy model
[@mackenzie2003estimating].

First, the model statement:

```{r print-file}
cat(paste(readLines('model.stan'), "\n", sep = ""), sep = "")
```

This model can be made faster by storing values for log(psi) and log(1 - psi), 
as done in [Bob Carpenter's single season example](https://groups.google.com/forum/#!topic/stan-users/ZuLmNl4Q9xc).

Fitting the model:

```{r fit-model, results='hide'}
library(reshape2)
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

simulate_data <- function(){
  nsite <- 100;
  nrep <- 3;     # repeat surveys
  nyear <- 10;  # nyears
  p <- 0.8;
  gamma <- .2
  phi <- .4
  psi1 <- .5

  Z <- array(dim=c(nsite, nyear))
  Z[, 1] <- rbinom(nsite, 1, psi1)
  for (t in 2:nyear){
    Z[, t] <- rbinom(nsite, 1, Z[, t - 1] * phi + (1 - Z[, t - 1]) * gamma)
  }

  Y <- array(dim=c(nsite, nyear, nrep))
  for (r in 1:nsite){
    for (t in 1:nyear){
      Y[r, t, ] <- rbinom(nrep, 1, Z[r, t] * p)
    }
  }
  return(list(nsite=nsite, nrep=nrep, nyear=nyear,
              p=p, gamma=gamma, phi=phi,
              psi1=psi1, Y=Y))
}

d <- simulate_data()

fit <- stan("model.stan", data = d)
```


```{r simulate-many-data-sets, results='hide'}
traceplot(fit, pars = c('gamma', 'phi', 'psi1'), ncol = 1)
```

Based on some initial exploration, this approach seems much much (much?) faster 
than explicitly modeling the latent occurrence states in JAGS, with better chain 
mixing and considerably less autocorrelation. 
